---
title: "CATFISH: Pathway Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CATFISH: Pathway Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction
CATFISH (MAGMA-based Gene and pathway Analysis with ACAT aggregation) provides a comprehensive toolkit for post-GWAS pathway enrichment analysis. It implements multiple p-value combination methods and combines them into an omnibus test with optional permutation-based calibration to account for gene-gene correlations.

## Installation

```{r installation}
# Install from GitHub (when available)
# devtools::install_github("nirwantandukar/CATFISH")

# Load the package
library(CATFISH)
```

## Workflow Overview

A typical CATFISH workflow consists of:

1. **Prepare gene location file** from GFF3 annotations
2. **Run MAGMA annotation** to map SNPs to genes
3. **Run MAGMA gene analysis** to get gene-level p-values
4. **Combine chromosome results** into a single file
5. **Optionally adjust p-values** for gene length/SNP density
6. **Run pathway tests** (individual or omnibus)

## Step 1: Create Gene Location File

Convert a GFF3 annotation file to MAGMA's gene location format:

```{r gff3-conversion}
# Convert GFF3 to MAGMA gene location format
loc <- gff3_to_geneloc(
  gff        = "path/to/annotation.gff3",
  out        = "inst/extdata/species.genes.loc",
  strict_chr = FALSE
)
```

## Step 2: MAGMA Annotation

Map SNPs to genes using MAGMA:
```{r magma-annotate}
# Set MAGMA path (required)
options(magma.path = "/path/to/magma")

# Run MAGMA annotation
ann <- magma_annotate(
  stats_file = "path/to/gwas_results.csv",
  rename_columns = c(
    CHR = "CHR",
    SNP = "SNP",
    POS = "Positions",
    PVALUE = "P-Value"
  ),
  gene_loc     = "inst/extdata/species.genes.loc",
  chr_map_path = "inst/extdata/species.genes.loc.chr_map.tsv",
  out_prefix   = "my_analysis",
  out_dir      = "annot",
  window       = c(10, 10),  # 10kb upstream/downstream
  sep          = ",",
  nonhuman     = TRUE
)
```

## Step 3: MAGMA Gene Analysis

Run MAGMA to compute gene-level statistics:

```{r magma-gene}
# Run MAGMA gene analysis per chromosome
magma_gene(
  bfile      = "path/to/genotype_file",
  gene_annot = ann$gene_annot,
  stats_file = "path/to/gwas_results.csv",
  sep        = ",",
  n_total    = 500,  # sample size
  rename_columns = c(CHR = "CHR", SNP = "SNP", PVALUE = "P-Value"),
  out_prefix = "my_analysis",
  out_dir    = "magma_genes_by_chr",
  gene_model = "multi=snp-wise",
  chroms     = 1:10,
  n_threads  = 10
)
```

## Step 4: Combine Chromosome Results

```{r combine-results}
# List all chromosome result files
files <- list.files(
  path = "magma_genes_by_chr",
  pattern = "^my_analysis_chr.*\\.genes\\.out$",
  full.names = TRUE
)

# Read and combine
gene_list <- lapply(files, function(f) {
  utils::read.table(f, header = TRUE, stringsAsFactors = FALSE)
})
genes_all <- do.call(rbind, gene_list)

# Deduplicate: keep smallest P per gene
genes_all <- genes_all[order(genes_all$GENE, genes_all$P), ]
genes_all <- genes_all[!duplicated(genes_all$GENE), ]
```

## Step 5: Optional P-value Adjustment

Adjust gene p-values for confounders like gene length and SNP density:

```{r adjust-p}
# Get gene lengths from GFF3
gene_lengths <- get_gene_lengths(
  gff3_file  = "path/to/annotation.gff3",
  output     = TRUE,
  output_dir = "inst/extdata",
  file_name  = "gene_lengths.tsv"
)

# Adjust p-values
adj_out <- catfish_adjust_gene_p(
  gene_results = genes_all,
  gene_lengths = gene_lengths,
  gene_col     = "GENE",
  nsnp_col     = "NSNPS",
  p_col        = "P",
  z_col        = "ZSTAT",
  len_gene_col = "gene_id",
  len_col      = "length"
)

# Extract adjusted results
genes_adj <- adj_out[, c("GENE", "ZSTAT", "P")]
```

## Step 6: Individual Pathway Tests

### ACAT (Aggregated Cauchy Association Test)

```{r acat-test}
acat_res <- catfish_acat_pathways(

  gene_results = genes_adj,
  species      = "maize",  # or "arabidopsis", "sorghum", "plant"
  gene_col     = "GENE",
  p_col        = "P",
  min_p        = 1e-15,
  output       = TRUE,
  out_dir      = "acat_results"
)
head(acat_res)
```

### Fisher's Method

```{r fisher-test}
fisher_res <- catfish_fisher_pathways(
  gene_results = genes_adj,
  species      = "maize",
  gene_col     = "GENE",
  p_col        = "P",
  min_p        = 1e-15,
  output       = TRUE,
  out_dir      = "catfish_fisher"
)
```

### Stouffer's Z Method

```{r stouffer-test}
stouffer_res <- catfish_stoufferZ_pathways(
  gene_results = genes_adj,
  species      = "maize",
  gene_col     = "GENE",
  z_col        = "ZSTAT",
  weight_col   = NULL,
  alternative  = "greater",
  output       = TRUE,
  out_dir      = "catfish_stouffer"
)
```

### Minimum P-value (minP)

```{r minp-test}
minp_res <- catfish_minp_pathways(
  gene_results = genes_adj,
  species      = "maize",
  gene_col     = "GENE",
  p_col        = "P",
  min_p        = 1e-15,
  do_fix       = TRUE,
  output       = TRUE,
  out_dir      = "catfish_minp"
)
```

### Adaptive Soft TFisher

```{r tfisher-test}
tfisher_res <- catfish_soft_tfisher_adaptive_pathways(

  gene_results = genes_adj,
  species      = "maize",
  gene_col     = "GENE",
  p_col        = "P",
  tau_grid     = c(0.01, 0.05, 0.1, 0.2),
  do_fix       = TRUE,
  output       = TRUE,
  out_dir      = "catfish_tfisher"
)
```

## Step 7: Omnibus Test (Recommended)

The omnibus test combines all five methods with permutation-based calibration:

```{r omnibus-test}
omni_res <- catfish_omni2_pathways(
  gene_results = genes_adj,
  species      = "maize",
  pmn_gene_col = "Gene-name",
  gene_col     = "GENE",
  p_raw_col    = "P",
  z_col        = "ZSTAT",

  # Test parameters
  tau_grid     = c(0.1, 0.05, 0.02, 0.01, 0.005, 0.001),
  min_p        = 1e-15,
  do_fix       = TRUE,
  stouffer_alternative = "greater",

  # Omnibus combination method
  omnibus      = "ACAT",  # or "minP"

  # Permutation calibration
  B_perm       = 10000L,
  perm_mode    = "mvn",   # "global", "mvn", "both", or "none"
  magma_cor_file = "path/to/gene_correlations.txt",
  make_PD      = TRUE,
  mvn_marginal = "uniform",

  # MVN calibration strategy (see below)
  mvn_calibrate_components = FALSE,

  seed         = 123,
  output       = TRUE,
  out_dir      = "catfish_omnibus_results"
)

head(omni_res)
```

### MVN Calibration Strategies

There are two ways to apply MVN-based LD correction, controlled by `mvn_calibrate_components`:

**Strategy 1: Calibrate omnibus only (`mvn_calibrate_components = FALSE`, default)**

1. Compute all 5 component test p-values analytically on observed data
2. Combine them into a single omnibus p-value (analytic)
3. Simulate null gene p-values from MVN(0, R_S) where R_S is the gene-gene
   correlation matrix for the pathway
4. For each null replicate, compute all 5 component tests and combine into
   a null omnibus p-value
5. Calibrate: compare observed omnibus against the null omnibus distribution

This is faster and appropriate when the main concern is the final omnibus
result. The component p-values (`acat_p`, `fisher_p`, etc.) remain analytic
(not LD-corrected).

**Strategy 2: Calibrate each component first (`mvn_calibrate_components = TRUE`)**

1. Compute all 5 component test p-values analytically on observed data
2. Simulate null gene p-values from MVN(0, R_S) for the pathway
3. For each component, calibrate its observed p-value against its own null
   distribution (e.g., observed ACAT p vs null ACAT p-values)
4. Combine the 5 MVN-calibrated component p-values into the omnibus
5. Calibrate that omnibus against a null omnibus built from per-draw
   calibrated components

This is more rigorous because each combination method has different
sensitivity to gene-gene correlation. For example, Fisher's method is
more affected by LD than ACAT. Calibrating components individually
removes method-specific LD bias before combining.

**Output columns by strategy:**

| Column | Strategy 1 | Strategy 2 |
|--------|-----------|-----------|
| `acat_p`, `fisher_p`, etc. | Analytic (uncorrected) | Analytic (uncorrected) |
| `acat_p_mvn_cal`, `fisher_p_mvn_cal`, etc. | `NA` | MVN-calibrated per component |
| `omni_p_mvn` | Calibrated omnibus | Calibrated omnibus (from calibrated components) |

**Recommendation:** Use Strategy 2 (`mvn_calibrate_components = TRUE`) for
publication-quality results. Use Strategy 1 for exploratory analysis or
when computation time is a concern.

```{r omnibus-strategy2, eval=FALSE}
# Strategy 2: Calibrate components individually (more rigorous)
omni_res_s2 <- catfish_omni2_pathways(
  gene_results = genes_adj,
  species      = "maize",
  gene_col     = "GENE",
  p_raw_col    = "P",
  z_col        = "ZSTAT",
  omnibus      = "ACAT",
  B_perm       = 10000L,
  perm_mode    = "mvn",
  magma_cor_file = "path/to/gene_correlations.txt",
  mvn_calibrate_components = TRUE,  # <<< KEY DIFFERENCE
  seed         = 123
)

# Now you get MVN-calibrated component p-values
head(omni_res_s2[, c("pathway_id", "acat_p", "acat_p_mvn_cal",
                       "fisher_p", "fisher_p_mvn_cal",
                       "omni_p_mvn")])
```

## Using Custom Pathways

Instead of built-in PMN pathways, you can provide custom pathways:

```{r custom-pathways}
# As a data.frame
custom_pw <- data.frame(
  pathway_id   = c("PWY1", "PWY1", "PWY2", "PWY2", "PWY2"),
  pathway_name = c("Pathway 1", "Pathway 1", "Pathway 2", "Pathway 2", "Pathway 2"),
  gene_id      = c("gene1", "gene2", "gene3", "gene4", "gene5")
)

# Or as a named list
custom_pw_list <- list(
  PWY1 = c("gene1", "gene2"),
  PWY2 = c("gene3", "gene4", "gene5")
)

# Use with any test
acat_custom <- catfish_acat_pathways(
  gene_results = genes_adj,
  pathways     = custom_pw,  # use custom instead of species
  gene_col     = "GENE",
  p_col        = "P"
)
```

## Output Interpretation

All functions return a data.frame with:

- `pathway_id`: Pathway identifier
- `pathway_name`: Human-readable pathway name
- `n_genes`: Number of genes in the pathway with valid p-values
- `gene_names`: Semicolon-separated list of gene IDs used
- Test-specific statistics and p-values

For the omnibus test, additional columns include:

**Component p-values (analytic):**
- `acat_p`, `fisher_p`, `minp_p_analytic`, `stouffer_p_analytic`, `tfisher_p_analytic`

**MVN-calibrated component p-values (only when `mvn_calibrate_components = TRUE`):**
- `acat_p_mvn_cal`, `fisher_p_mvn_cal`, `minp_p_mvn_cal`, `stouffer_p_mvn_cal`, `tfisher_p_mvn_cal`

**Omnibus p-values:**
- `omni_p_analytic`: Analytic omnibus (no LD correction)
- `omni_p_global`: Global permutation-calibrated
- `omni_p_mvn`: MVN-calibrated (accounts for gene-gene LD)
- `omni_p_final`: Best available (MVN > global > analytic)

**Multiple testing corrections (applied to `omni_p_final`):**
- `omni_p_final_BH`: Benjamini-Hochberg FDR
- `omni_p_final_q_storey`: Storey's q-value
- `omni_p_final_q_bky10`: Benjamini-Krieger-Yekutieli at q=0.10
- `omni_p_final_q_ihw10`: IHW using pathway size as covariate

**Diagnostics:**
- `dominant_component`: Which method gave the smallest p-value
- `agreement_score`: Fraction of methods with p < 0.05
- `agreement_k` / `agreement_m`: Count of significant / available methods
- `calibration_impact`: How calibration changed significance status
- `genes_used`: Semicolon-separated gene IDs in the pathway
- `gene_pvals_used`: Semicolon-separated gene p-values

## Best Practices

1. **Always use permutation calibration** (`B_perm >= 10000`) for publication-quality results
2. **Use MVN mode** (`perm_mode = "mvn"`) when genes are correlated (typical in GWAS)
3. **Adjust for gene length** if you observe correlation between gene size and significance
4. **Use the omnibus test** rather than cherry-picking individual tests
5. **Apply FDR correction** to the final p-values for multiple testing

## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
